A \proglang{Python}/\proglang{Zig} optimized and customizable implementation for the $\rho_{DCCA}$ and $DMC_x^2$ methods

### 2025-01-29 paper_v02/article.tex
paper version 02

%% - \Abstract{} almost as usual
\Abstract{
  This paper presents tha \pkg{Zebende}, a \proglang{Python} package written in \proglang{Python} and \proglang{Zig}, that calculates the $DFA$, $DCCA$ $\rho_{DCCA}$ and the $DMC_x^2$. The package presents an optimized algorithm that significantly improves the calculations speed. A comparison with other packages that calculates the .The package is also the first to implement the $DMC_x^2$ coefficient for any number of series.
}

@@

The Detrended Cross-correlation Coefficient ($\rho_{DCCA}$)~\citep{Zebende2011} is a widely used coefficient that measures the cross-correlation between tow non-stationary time series. It´s an extension of the Detrended Fluctuation Analysis ($DFA$)~\citep{Peng_1994} and the Detrended Cross-correlation Analysis ($DCCA$)~\citep{Podobnik2008}: while the $DFA$ calculates the self-affinity and long-memory properties of a time series data, and the $DCCA$ analyses power-law cross correlations between two different non-stationarity time series, the $\rho_{DCCA}$ coefficient quantifies this cross-correlation in simple values ranging from $-1$ to $1$, where $-1$ indicates a perfect anti-correlation between the series, $1$ a perfect correlation and zero ($0$) no correlation at all.

@@

\begin{enumerate}
  \item Enhance performance;
  \item avoid redundant calculations;
  \item make the outputs compatibles with other data analyses tools (including data manipulation, machine learn and statistical packages);
  \item create a customizable and modular set of tools;
  \item facilitate package evolution and maintenance;
  \item deliver an easy to use package.
\end{enumerate}

@@

The first draft of the code was written in pure \proglang{Python}, to rapidly prototype the way users will interact with the package. Figures \ref{fig:pdcca_chart} and \ref{fig:dmc_chart} presents simplified flowcharts illustrating how to use the package and how the main functions ($\rho_{DCCA}$ and $DMA_x^2$) works.

@@

The first two inputs are the same for functions \verb"p_dcca()" and \verb"dmcx2()": \verb"input_data" receive the integrated series and the \verb"tws" receives an 1D array representing the time scales (box size) described in the algorithms on Section \ref{sec:calculations}. The \verb"input_data" is a 2D array of 64 bits floating point data. The \verb"tws" accepts integers and, for convenience reasons, also floating points. Since the size of the boxes needs to be integers, in case of floating points, the values will be converted to integers by ignoring the decimal values(truncating). This two inputs are colored in light gray in Figures \ref{fig:pdcca_chart} and \ref{eq:dmcx2}, indicating mandatory inputs.

@@

It´s important to understand the calculation steps, the role of the \verb"DCCA_of" array and how it fits in the goals of the package implementation. The code below is part of the \proglang{Python} implementation of the $\rho_{DCCA}$ function,

@@

The first \verb"for" loop in the code operates over the values of the \verb"tws" input, asserting that the step 6 of the $DFA$ and $DCCA$ methods in Section \ref{sec:calculations}. Three temporary arrays are allocated for each time scale. The first, \verb"f2dfa_n", to store the calculations of the step 4 of the $DFA$. The number of lines of this array correspond to the number of boxes in the current time scale ($N - n$) and the columns is the number of series in the analysis. Array  \verb"dcca_n", holds the values calculated in the step 4 of the $DCCA$. The number of lines also correspond to the number of boxes but the number of columns equals the number of pairs (rows) in the \verb"DCCA_of" input. The \verb"detrended_mat" array has a number of lines is the number of points in a time scale box ($n +1$) and column count also equal to the number of time series. This last array is very important to the implementation goals.

In each box, the \verb"detrended_series" function execute a one degree polynomial fit, subtract the values of the serie with the value given by the interpolated curve and stores this values in the \verb"detrended_mat". After the $DFA$ is calculated. With all the detrended values calculated, another loop, for all the rows in the \verb"DCCA_of", uses the \verb"detrended_mat" and the $DFA$ arrays to calculate the $DCCA$. This approach avoid redundant calculation. If the \verb"DCCA_of" is $[[0,1], [0,2], [1,2]]$, the detrended and $DFA$ values for each serie are calculated just once. If in a set of four series, a research intend to calculate the $DCCA$ between the first serie and all the others along with the two last series, inputting  \verb"DCCA_of" $= [[0,1], [0,2], [0,3], [2,3]]$. In this case, the \verb"detrended_mat" will be calculated just once for each series and the $DCCA$ between series $[[1,2], [1,3]]$ will be skipped, revealing a very customizable algorithm.

The function named \verb"P_DCCA_output_function" in the code above,is a pointer to other functions. One that outputs the  $\rho_{DCCA}$ results in the form of a table (rows for each time scale and columns for each \verb"DCCA_of" pair) and the other outputs it the form of a 3D matrix, where each level is the matrix in Equation \ref{eq:p_dcca_matrix} for one of the time scales. This behavior is driven by the \verb"P_DCCA_output_matrix" input (represented as the light green node in Figure \ref{fig:pdcca_chart}), where \verb"False" means table output and \verb"True" matrix output. This is very convenient for calculating the $DMC_x^2$. There are two utility functions to transform a table output in a matrix one (\verb"p_dcca_simple_to_matrix") and also the other way around (\verb"p_dcca_matrix_to_simple").

The \verb"dmcx2" function runs the \verb"p_dcca()" with \verb"P_DCCA_output_matrix" set to \verb"True". There is no \verb"DCCA_of" input for the $DMC_x^2$ function, instead there is a \verb"dmcx2_of" parameter. This input receives a 2D matrix where each line represents the indexes of the series to be used in Equation \ref{eq:dmcx2}. The first value of each row is the index of the serie used as the dependent variable, the others, the independent ones. There are two literal strings that can be used, for convenience as inputs for this parameter: \verb"'all-full'", that generates a 2D array with every series as the dependent variable against all the others; and \verb"'first-full'", with only one row, having the index zero series as the dependente variable in relation with all the others. The \verb"'all-full'" option is operated by the function \verb"dmc_of_all_as_y".

There is a restriction to the \verb"dmcx2_of" array: the independent values must be in crescent order. There is a check in the \verb"dmcx2" that raises an error if the values are out of order, and suggests the use of a utility function named \verb"ordering_x_dmcx2_of". The code below shows an example of a manually defined \verb"dmcx2_of" array with 2 elements out of order in the second row, the \verb"AssertionError" output, the way of fixing it with the \verb"ordering_x_dmcx2_of" function and finally the outputted ordered array.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
  This paper presents tha \pkg{Zebende}, a \proglang{Python} package written in \proglang{Python} and \proglang{Zig}, that calculates the $DFA$, $DCCA$ $\rho_{DCCA}$ and the $DMC_x^2$. The package presents an optimized algorithm that significantly improves the calculations speed. A comparison with other packages that calculates the $\rho_{DCCA}$. The package is also the first to implement the $DMC_x^2$ coefficient in a package and the first algorithm to calculates it for any number of time series.
}

@@

The Detrended Cross-correlation Coefficient ($\rho_{DCCA}$)~\citep{Zebende2011} is a widely used coefficient that measures the cross-correlation between two non-stationary time series. It´s an extension of the Detrended Fluctuation Analysis ($DFA$)~\citep{Peng_1994} and the Detrended Cross-correlation Analysis ($DCCA$)~\citep{Podobnik2008}: while the $DFA$ calculates the self-affinity and long-memory properties of a time series data, and the $DCCA$ analyses power-law cross correlations between two different non-stationarity time series, the $\rho_{DCCA}$ coefficient quantifies this cross-correlation in simple values ranging from $-1$ to $1$, where $-1$ indicates a perfect anti-correlation between the series, $1$ a perfect correlation and zero ($0$) no correlation at all.

@@

\begin{enumerate}
  \item Enhance performance;
  \item avoid redundant calculations;
  \item make the outputs compatibles with other data analyses tools (including data manipulation, machine learn and statistical packages);
  \item manage multiple time series inputs;
  \item operate the $DMC_x^2$ for any number of series;
  \item create a customizable and modular set of tools;
  \item facilitate package evolution and maintenance;
  \item deliver an easy to use package.
\end{enumerate}

@@

The first draft of the code was written in pure \proglang{Python}, to rapidly prototype the way users will interact with the package. Figures \ref{fig:pdcca_chart} and \ref{fig:dmc_chart} presents simplified flowcharts illustrating how to use the package and how the main functions ($\rho_{DCCA}$ and $DMC_x^2$) works.

@@

The first two inputs are the same for functions \verb"p_dcca()" and \verb"dmcx2()": \verb"input_data" receive the integrated series and the \verb"tws" receives an 1D array representing the time scales (box size) described in the algorithms on Section \ref{sec:calculations}. The \verb"input_data" is a 2D array of 64 bits floating point data. The \verb"tws" accepts integers and, for convenience reasons, also floating points. Since the size of the boxes needs to be integers, in case of floating points, the values will be converted to integers by ignoring the decimal values (truncating). This two inputs are colored in light gray in Figures \ref{fig:pdcca_chart} and \ref{eq:dmcx2}, indicating mandatory inputs.

@@

It´s important to understand the calculation steps, the role of the \verb"DCCA_of" array and how it fits in the goals of the package implementation. The code below is part of the pure \proglang{Python} implementation of the $\rho_{DCCA}$ function,

@@

The first \verb"for" loop in the code operates over the values of the \verb"tws" input, ensuring that step 6 of the $DFA$ and $DCCA$ methods, presented in Section \ref{sec:calculations}, is being carried out. In other words, the calculations will be applied, sequentially, to every single value in the \verb"tws" array. Three temporary arrays are allocated and resized for each time scale. The first, \verb"f2dfa_n", is used to store the calculations of the step 4 of the $DFA$~($f_{DFA}^{2}$). The number of lines of this array correspond to the number of boxes in the current time scale ($N - n$, resized for each time scale $n$) and the columns is the number of series in the analysis (same size for every value of $n$). Second, the array  \verb"dcca_n" holds the values calculated in the step 4 of the $DCCA$~(to calculate $f_{DCCA}^{2}(n, i)$). The number of lines also correspond to the number of boxes(resized for each $n$) but the number of columns equals the number of pairs (rows) in the \verb"DCCA_of" input (same size in each $n$). The \verb"detrended_mat" array has a number of lines equal to the number of points in a time scale box ($n +1$, resized for each $n$) and column count also equal to the number of time series (same size in each $n$). The first two tempoaray arrays will store data for all the boxes in the time scale, the last one will be used in each box and will have the values replaced in the next one, until the last box of the time scale. Than all the arrays will be cleared and recreated with shapes calculated with the new value of $n$.

After the allocation of the temporary arrays, the second \verb"for" loop operates in every box for a certain time scale $n$. In each box, the \verb"detrended_series()" function execute a one degree polynomial fit, subtract the values of the serie with the value given by the interpolated curve($DV$) and stores this values in the \verb"detrended_mat". After that, the \textbf{mean of the square of each $DV$ in the box} ($f_{DFA}^{2}(n, i)$) is calculated in the current box for every time series and stored in the \verb"f2dfa_n" array in the line associated with the current box, and the column related to each time series.

The next nested \verb"for" loop operates over the \verb"DCCA_of" array. For each line (pair) in the \verb"DCCA_of" array, the $f_{DCCA}^{2}(n, i)$ is evaluated by multiplying the corresponding $DV$ from the two series  boxes in the current pair, getting the mean, and stores it in the \verb"dcca_n" temporary array.

After all the pairs are calculated, the algorithm goes back to the

The function named \verb"P_DCCA_output_function()" in the code above,is a pointer to other functions. One that outputs the  $\rho_{DCCA}$ results in the form of a table (rows for each time scale and columns for each \verb"DCCA_of" pair) and the other outputs it the form of a 3D matrix, where each level is the matrix in Equation \ref{eq:p_dcca_matrix} for one of the time scales. This behavior is driven by the \verb"P_DCCA_output_matrix" input (represented as the light green node in Figure \ref{fig:pdcca_chart}), where \verb"False" means table output and \verb"True" matrix output. This is very convenient for calculating the $DMC_x^2$. There are two utility functions to transform a table output in a matrix one (\verb"p_dcca_simple_to_matrix()") and also the other way around (\verb"p_dcca_matrix_to_simple()").

The \verb"dmcx2" function runs the \verb"p_dcca()" with \verb"P_DCCA_output_matrix" set to \verb"True". There is no \verb"DCCA_of" input for the $DMC_x^2$ function, instead there is a \verb"dmcx2_of" parameter. This input receives a 2D matrix where each line represents the indexes of the series to be used in Equation \ref{eq:dmcx2}. The first value of each row is the index of the serie used as the dependent variable, the others, the independent ones. There are two literal strings that can be used, for convenience as inputs for this parameter: \verb"'all-full'", that generates a 2D array with every series as the dependent variable against all the others; and \verb"'first-full'", with only one row, having the index zero series as the dependente variable in relation with all the others. The \verb"'all-full'" option is operated by the function \verb"dmc_of_all_as_y()".

There is a restriction to the \verb"dmcx2_of" array: the independent values must be in crescent order. There is a check in the \verb"dmcx2" that raises an error if the values are out of order, and suggests the use of a utility function named \verb"ordering_x_dmcx2_of()". The code below shows an example of a manually defined \verb"dmcx2_of" array with 2 elements out of order in the second row, the \verb"AssertionError" output, the way of fixing it with the \verb"ordering_x_dmcx2_of()" function and finally the outputted ordered array.
